## Student performance (incl. alcohol consumption)

This is an exercise on data from two questionnaires related to student performance (especially including alcohol consumption). More information about the data sets  can be found in the link below.

<https://archive.ics.uci.edu/ml/datasets/Student+Performance>

The data have been adjusted as follow :  
* The variables not used for joining the two data have been combined by averaging (including the grade variables)  
* 'alc_use' is the average of 'Dalc' and 'Walc'  
* 'high_use' is TRUE if 'alc_use' is higher than 2 and FALSE otherwise  

The purpose of the analysis is to study the relationships between high/low alcohol consumption and some of the other variables in the data. To do this, 4 interesting variables have been chosen. The hypothesis for the chosen variables and  about their relationships with alcohol consumption are as follows:
Variables 1-4 (listed below) with values hig, A, high and no will increase alcohol consumption

1.goout - going out with friends (numeric: from 1 - very low to 5 - very high):high  
2.Pstatus - parent's cohabitation status (binary: 'T' - living together or 'A' - apart):A  
3.freetime - free time after school (numeric: from 1 - very low to 5 - very high):high  
4.activities - extra-curricular activities (binary: yes or no):no  
  
  The distribution of chosen variables and relationships with alcohol consumption is explored with cross-tabulations, bar plot and box plots.  
```{r}

```
  
The findings of the exploration and the hypothesis seems...  

The logistic regression is used explore the relationship between the chosen variables and the binary high/low alcohol consumption variable as the target variable. The summary of the fitted model is as follow:  
```{r}

```
Further  the coefficients of the model as odds ratios and provide confidence intervals for them:  
Interpretation of the results and comparison them to previously stated hypothesis.  

The variables which, according to your logistic regression model, had a statistical relationship with high/low alcohol consumption, explore the predictive power of the model. Here is provided a 2x2 cross tabulation of predictions versus the actual values and  displayed a graphic visualizing both the actual values and the predictions.
```{r}

```

The total proportion of inaccurately classified individuals (= the training error)  is computed and comments on all the results can be find here:

Comparison of the performance of the model with performance achieved by some simple guessing strategy. 

Further 10-fold cross-validation on the model.  
```{r}

```


Does the model have better test set performance (smaller prediction error using 10-fold cross-validation) compared to the model introduced in DataCamp (which had about 0.26 error). Could you find such a model?  


The cross-validation to compare the performance of different logistic regression models (= different sets of predictors). The task has been started with a very high number of predictors and explored the changes in the training and testing errors as went to models with less predictors.   
```{r}

```

Drawed a graph to display the trends of both training and testing errors by the number of predictors in the model.   
```{r}

```

